---
title: 'Reproducible Research: Peer Assessment 1'
author: "edgarodriguez"
date: "7/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache = FALSE,fig.width = 12,fig.height = 4)
```

## Loading and preprocessing the data
```{r} 
# Unzipping and reading *.csv file
path = getwd()
unzip("activity.zip", exdir = path)
activity <- read.csv("activity.csv")

# Viewing the processed data
summary(activity)
```
## What is mean total number of steps taken per day?

```{r}
library(ggplot2)
activityTotalSteps <- with(activity, aggregate(steps, by = list(date), sum, na.rm = TRUE))
names(activityTotalSteps) <- c("Date", "Steps")

totalStepsdf <- data.frame(activityTotalSteps)
g <- ggplot(totalStepsdf, aes(x = Steps)) + 
    geom_histogram(breaks = seq(0, 25000, by = 1000), fill = "#335252", col="#2D3033") + 
   
    xlab("Total Steps") + 
    ylab("Number of days") + 
    ggtitle("Total of Steps Taken by Day")+
    geom_vline(aes(xintercept=mean(Steps)),color="#AA4B41", linetype="dashed", size=1, show.legend=TRUE)+
    geom_vline(aes(xintercept=median(Steps)),color="#F9BA32", linetype="dashed", size=1)
               
print(g)
```
```{r, echo=FALSE} 
meanSteps <-mean(activityTotalSteps$Steps)
medianSteps <-median(activityTotalSteps$Steps)
```
Mean = `r meanSteps` "Red line", Median = `r medianSteps` "Yellow"

## What is the average daily activity pattern?
```{r}
# Calculating the average number of steps taken, averaged across all days by 5-min intervals.
averageDailyActivity <- aggregate(activity$steps, by = list(activity$interval), 
                                  FUN = mean, na.rm = TRUE)


# Changing col names
names(averageDailyActivity) <- c("Interval", "Mean")

# Converting the data set into a dataframe
averageActivitydf <- data.frame(averageDailyActivity)

# Plotting on ggplot2
f <- ggplot(averageActivitydf, mapping = aes(Interval, Mean)) + 
    geom_line(col = "#335252") +
    xlab("Day period (Intervals of 5 minutes)") + 
    ylab("Average Number of Steps") + 
    ggtitle("Average Number of Steps Per Interval")
print(f)
```

### The 5-minute interval that, on average, contains the maximum number of steps:
```{r}
maxInterval <- averageDailyActivity[which.max(averageDailyActivity$Mean), ]$Interval
print(maxInterval)
```
Which means: `r maxInterval%/%60`:`r maxInterval%%60` hrs

## Imputing missing values
```{r}
sum(is.na(activity$steps))
```

### Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r}
imputedSteps <- averageDailyActivity$Mean[match(activity$interval, averageDailyActivity$Interval)]
```

### Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r}
activityImputed <- transform(activity, 
                             steps = ifelse(is.na(activity$steps), yes = imputedSteps, no = activity$steps))

totalActivityImputed <- aggregate(steps ~ date, activityImputed, sum)

names(totalActivityImputed) <- c("date", "Steps")
```

### Histogram of the total number of steps taken each day after missing values are imputed

```{r}
totalImputedStepsdf <- data.frame(totalActivityImputed)
g <- ggplot(totalImputedStepsdf, aes(x = Steps)) + 
    geom_histogram(breaks = seq(0, 25000, by = 1000), fill = "#335252", col="#2D3033") + 
    
    xlab("Total Steps") + 
    ylab("Number of days") + 
    ggtitle("Total of Steps Taken by Day")+
    geom_vline(aes(xintercept=mean(Steps)),color="#AA4B41", linetype="dashed", size=1, show.legend=TRUE)+
    geom_vline(aes(xintercept=median(Steps)),color="#F9BA32", linetype="dashed", size=1)

print(g)
```
## Are there differences in activity patterns between weekdays and weekends?

```{r}
activity$date <- as.Date(strptime(activity$date, format="%Y-%m-%d"))

activity$dayType <- sapply(activity$date, function(x) {
    if(weekdays(x) == "Saturday" | weekdays(x) == "Sunday")
    {y <- "Weekend"}
    else {y <- "Weekday"}
    y
})

```
### Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r}

activityByDay <-  aggregate(steps ~ interval + dayType, activity, mean, na.rm = TRUE)

Plot <-  ggplot(activityByDay, aes(x = interval , y = steps, color = dayType)) + 
    geom_line() + ggtitle("Average Daily Steps by Day Type") + 
    xlab("Interval") + 
    ylab("Average Number of Steps") +
    facet_wrap(~dayType, ncol = 1, nrow=2) +
    scale_color_discrete(name = "Day Type")
print(Plot)
```
### Code could be found in Script.R file